Storm surge modelling in the Strait of Georgia

### 2014-12-03 RoughWork/manuscript.tex
Merged tables into one

The maximum water level and maximum residual are important factors in storm surge hindcasts but a better indicator of overall model performance is provided by a group of statistical measures, including the mean absolute error, the observed mean, the modelled mean, $\gamma^2$, and the Willmott skill score for both the water level and residuals (Table \ref{tab:feb2006stat}). The statistics are calculated with hourly output over the three-day period displayed in Figure \ref{fig:feb2006}. To begin with, the mean absolute error in the water level ranges from 7.6 cm at Victoria to 13.1 cm at Campbell River, whereas, the mean absolute error in the residuals is smaller at all stations, however, this measure does not account for the large variability in water levels due to the tides. A statistic that does take into account this variability is $\gamma^2$ which suggests very good agreement in both the predicted water levels and residuals. Good agreement is also reflected in high Willmott skill scores at all locations. Note that the $\gamma^2$ measures and Willmott skill scores degrade slightly for the residuals, likely because the well-tuned tidal signal dominates the water level predictions.

\begin{table}[h]
\centering 
\begin{tabular}{|l |c c c c c | c c c c c|} 
\hline 
& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
\hline 
Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
\hline 
Point Atkinson & 10.5           & 3.46                 & 3.54                 &   0.012    & 0.996  & 6.2            &  0.273                 & 0.309                  & 0.065     & 0.979     \\
Victoria       &  7.6           & 2.18                 & 2.19                 &   0.032    & 0.992  & 6.1            &  0.235                 & 0.278                  & 0.085     & 0.970     \\
Patricia Bay   &  9.1           & 2.60                 & 2.64                 &   0.024    & 0.992  & 6.8            &  0.223                 & 0.278                  & 0.070     & 0.969     \\
Campbell River & 13.1           & 3.20                 & 3.21                 &   0.041    & 0.990  & 5.8            &  0.303                 & 0.300                  & 0.079     & 0.978     \\
\hline 
\end{tabular}
\caption{Evaluation of model performance for the Feb 2006 simulation. Mean absolute error ($\bar{e}$), observed mean (subscript o), modelled mean (subscript m), $\gamma^2$, and the Willmott skill score ($WS$) are provided for water level predictions ($\eta$) and residuals ($\delta$) during the period displayed in Figure \ref{fig:feb2006} from Feb 3, 2006 to Feb 6, 2006.}
\label{tab:feb2006stat} 
\end{table} 
%anything else to include? statistics calculated in Feb 2006 -paper production.ipynb and saved in statistics_feb2006.CV 
%I also have standard deviations, correlations, ...

@@

An additional storm surge hindcast for Nov 2006 is provided as another means of assessing model performance and is mentioned briefly here. On Nov 16, 2006 communities in the Strait of Georgia experienced a strong wind storm with speeds upwards of 20 m s$^{-1}$ near Point Atkinson resulting in observed water levels of 5.1 m and a maximum surge amplitude of 83.7 cm. The surge amplitude was lower at the other stations, reaching no more than 68 cm. The model performs reasonably well in this hindcast; Willmott skill scores are greater than 0.94 (Table \ref{tab:nov2006stat}).  The best match in maximum surge occurred at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and too early at the other the stations.

\begin{table}[h]
\centering 
\begin{tabular}{|l |c c c c c | c c c c c|} 
\hline 
& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
\hline 
Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
\hline 
Point Atkinson &  9.8           &  3.42                & 3.43                 &   0.034    & 0.991  &  7.7           &  0.290                 & 0.247                  &  0.128     & 0.950 \\
Victoria       &  9.1           &  2.16                & 2.11                 &   0.070    & 0.978  &  6.2           &  0.222                 & 0.202                  &  0.171     & 0.946 \\
Patricia Bay   & 10.0           &  2.57                & 2.53                 &   0.049    & 0.985  &  6.6           &  0.235                 & 0.203                  &  0.144     & 0.947 \\
Campbell River & 10.7           &  3.20                & 3.26                 &   0.037    & 0.989  &  6.2           &  0.254                 & 0.236                  &  0.140     & 0.952 \\
\hline 
\end{tabular}
\caption{Evaluation of model performance for the Nov 2006 hindcast. Statistics were calculated over the period Nov 14, 2006 to Nov 17, 2006.}
\label{tab:nov2006stat} 
\end{table}

@@

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 28 cm higher than the modelled residual, however the timing of the surge agrees almost exactly as displayed in Figure \ref{fig:dec2006}. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 44 cm. At all of the locations, the model predicted the maximum surge to occur within 3 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb 4, 2006 case. All Willmott skill scores rank above 0.89 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:dec2006stat}). The modelled residuals are on average too low and have a lower skill score than the Feb 2006 case. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

\begin{table}[h]
\centering 
\begin{tabular}{|l |c c c c c | c c c c c|} 
\hline 
& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
\hline 
Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
\hline 
Point Atkinson & 10.5           &  3.45                & 3.47                 &   0.025    & 0.994  & 11.1           &  0.212                 & 0.186                  &  0.275     & 0.891 \\
Victoria       &  9.6           &  2.23                & 2.16                 &   0.042    & 0.984  &  7.9           &  0.190                 & 0.156                  &  0.223     & 0.911 \\
Patricia Bay   &  9.2           &  2.63                & 2.58                 &   0.031    & 0.990  &  9.5           &  0.193                 & 0.149                  &  0.260     & 0.889 \\
Campbell River & 10.6           &  3.23                & 3.28                 &   0.030    & 0.992  &  9.4           &  0.191                 & 0.163                  &  0.247     & 0.905 \\
\hline 
\end{tabular}
\caption{Evaluation of model performance for the Dec 2006 hindcast. Statistics were calculated over the period Dec 14, 2006 to Dec 17, 2006.}
\label{tab:dec2006stat} 
\end{table}

@@

The modelled water levels and residuals are too high for this hindcast, except at Victoria (Table \ref{tab:nov2009stat}). Although Victoria has the smallest mean error in both the water level and residuals, its Willmott skill score and $\gamma^2$ values rank low when compared to the other stations. Generally, the skills scores are very good, scoring above 0.89 at all stations.

\begin{table}[h]
\centering 
\begin{tabular}{|l |c c c c c | c c c c c|} 
\hline 
& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
\hline 
Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
\hline 
Point Atkinson & 14.0           &  3.44                & 3.55                 &   0.012    & 0.995  &  6.1           &  0.313                 & 0.331                  &  0.237     & 0.914 \\
Victoria       &  9.9           &  2.26                & 2.24                 &   0.023    & 0.994  &  4.8           &  0.323                 & 0.293                  &  0.245     & 0.899 \\
Patricia Bay   & 11.2           &  2.60                & 2.65                 &   0.020    & 0.993  &  6.0           &  0.277                 & 0.290                  &  0.272     & 0.901 \\
Campbell River & 13.5           &  3.22                & 3.27                 &   0.019    & 0.995  &  4.9           &  0.326                 & 0.314                  &  0.217     & 0.924 \\
\hline 
\end{tabular}
\caption{Evaluation of model performance for the Nov 2009 hindcast. Statistics were generated over the period Nov 17, 2009 to Nov 22, 2009.}
\label{tab:nov2009stat} 
\end{table}

>>>>

The maximum water level and maximum residual are important factors in storm surge hindcasts but a better indicator of overall model performance is provided by a group of statistical measures, including the mean absolute error, the observed mean, the modelled mean, $\gamma^2$, and the Willmott skill score for both the water level and residuals (Table \ref{tab:statistics}). The statistics are calculated with hourly output over the three-day period displayed in Figure \ref{fig:feb2006}. To begin with, the mean absolute error in the water level ranges from 7.6 cm at Victoria to 13.1 cm at Campbell River, whereas, the mean absolute error in the residuals is smaller at all stations, however, this measure does not account for the large variability in water levels due to the tides. A statistic that does take into account this variability is $\gamma^2$ which suggests very good agreement in both the predicted water levels and residuals. Good agreement is also reflected in high Willmott skill scores at all locations. Note that the $\gamma^2$ measures and Willmott skill scores degrade slightly for the residuals, likely because the well-tuned tidal signal dominates the water level predictions.

\begin{table}[h]
\centering 
\begin{tabular}{|l |c c c c c | c c c c c|} 
\hline 
& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
\hline 
Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
\hline 
Feb 2006       &                &                      &                      &            &        &                &                        &                        &           &           \\
\hline 
Point Atkinson & 10.5           & 3.46                 & 3.54                 &   0.012    & 0.996  & 6.2            &  0.273                 & 0.309                  & 0.065     & 0.979     \\
Victoria       &  7.6           & 2.18                 & 2.19                 &   0.032    & 0.992  & 6.1            &  0.235                 & 0.278                  & 0.085     & 0.970     \\
Patricia Bay   &  9.1           & 2.60                 & 2.64                 &   0.024    & 0.992  & 6.8            &  0.223                 & 0.278                  & 0.070     & 0.969     \\
Campbell River & 13.1           & 3.20                 & 3.21                 &   0.041    & 0.990  & 5.8            &  0.303                 & 0.300                  & 0.079     & 0.978     \\
\hline 
Nov 2006       &                &                      &                      &            &        &                &                        &                        &           &           \\
\hline 
Point Atkinson &  9.8           &  3.42                & 3.43                 &   0.034    & 0.991  &  7.7           &  0.290                 & 0.247                  &  0.128     & 0.950 \\
Victoria       &  9.1           &  2.16                & 2.11                 &   0.070    & 0.978  &  6.2           &  0.222                 & 0.202                  &  0.171     & 0.946 \\
Patricia Bay   & 10.0           &  2.57                & 2.53                 &   0.049    & 0.985  &  6.6           &  0.235                 & 0.203                  &  0.144     & 0.947 \\
Campbell River & 10.7           &  3.20                & 3.26                 &   0.037    & 0.989  &  6.2           &  0.254                 & 0.236                  &  0.140     & 0.952 \\
\hline 
Dec 2006       &                &                      &                      &            &        &                &                        &                        &           &           \\
\hline 
Point Atkinson & 10.5           &  3.45                & 3.47                 &   0.025    & 0.994  & 11.1           &  0.212                 & 0.186                  &  0.275     & 0.891 \\
Victoria       &  9.6           &  2.23                & 2.16                 &   0.042    & 0.984  &  7.9           &  0.190                 & 0.156                  &  0.223     & 0.911 \\
Patricia Bay   &  9.2           &  2.63                & 2.58                 &   0.031    & 0.990  &  9.5           &  0.193                 & 0.149                  &  0.260     & 0.889 \\
Campbell River & 10.6           &  3.23                & 3.28                 &   0.030    & 0.992  &  9.4           &  0.191                 & 0.163                  &  0.247     & 0.905 \\
\hline  
Nov 2009       &                &                      &                      &            &        &                &                        &                        &           &           \\
\hline 
Point Atkinson & 14.0           &  3.44                & 3.55                 &   0.012    & 0.995  &  6.1           &  0.313                 & 0.331                  &  0.237     & 0.914 \\
Victoria       &  9.9           &  2.26                & 2.24                 &   0.023    & 0.994  &  4.8           &  0.323                 & 0.293                  &  0.245     & 0.899 \\
Patricia Bay   & 11.2           &  2.60                & 2.65                 &   0.020    & 0.993  &  6.0           &  0.277                 & 0.290                  &  0.272     & 0.901 \\
Campbell River & 13.5           &  3.22                & 3.27                 &   0.019    & 0.995  &  4.9           &  0.326                 & 0.314                  &  0.217     & 0.924 \\
\hline 
Dec 2012-CGRF &                &                      &                      &            &        &                &                        &                        &           &           \\
\hline 
Point Atkinson &  9.1           &  3.36                & 3.40                 &   0.006    & 0.998  &  5.0           &  0.300                 & 0.290                  &  0.161     & 0.949 \\
Victoria       & 14.6           &  2.15                & 2.08                 &   0.039    & 0.988  &  4.8           &  0.309                 & 0.274                  &  0.150     & 0.937 \\
Patricia Bay   & 14.2           &  2.55                & 2.51                 &   0.026    & 0.992  &  4.7           &  0.300                 & 0.274                  &  0.140     & 0.945 \\
Campbell River & 11.6           &  3.06                & 3.09                 &   0.013    & 0.996  &  5.7           &  0.241                 & 0.278                  &  0.182     & 0.921 \\
\hline 
Dec 2012-HRDPS &                &                      &                      &            &        &                &                        &                        &           &           \\
\hline 
Point Atkinson &  8.8           &  3.35                & 3.39                 &   0.006    & 0.998  &  5.4           &  0.300                 & 0.278                  &  0.163     & 0.943 \\
Victoria       & 14.3           &  2.15                & 2.09                 &   0.038    & 0.988  &  4.6           &  0.309                 & 0.277                  &  0.145     & 0.941 \\
Patricia Bay   & 14.0           &  2.55                & 2.52                 &   0.025    & 0.993  &  4.5           &  0.300                 & 0.278                  &  0.130     & 0.953 \\
Campbell River & 11.6           &  3.06                & 3.09                 &   0.014    & 0.993  &  5.6           &  0.241                 & 0.275                  &  0.183     & 0.924 \\
\hline 
\end{tabular}
\caption{Evaluation of model performance for all hindcasts presented. Mean absolute error ($\bar{e}$), observed mean (subscript o), modelled mean (subscript m), $\gamma^2$, and the Willmott skill score ($WS$) are provided for water level predictions ($\eta$) and residuals ($\delta$). Statisitics are calculated over the time period displayed in esch simulation's corresponding figure.}
\label{tab:statistics} 
\end{table} 
%anything else to include? statistics calculated in Feb 2006 -paper production.ipynb and saved in statistics_feb2006.CV 
%I also have standard deviations, correlations, ...

@@

An additional storm surge hindcast for Nov 2006 is provided as another means of assessing model performance and is mentioned briefly here. On Nov 16, 2006 communities in the Strait of Georgia experienced a strong wind storm with speeds upwards of 20 m s$^{-1}$ near Point Atkinson resulting in observed water levels of 5.1 m and a maximum surge amplitude of 83.7 cm. The surge amplitude was lower at the other stations, reaching no more than 68 cm. The model performs reasonably well in this hindcast; Willmott skill scores are greater than 0.94 (Table \ref{tab:statistics}).  The best match in maximum surge occurred at Campbell River with a 6.2 cm underestimate by the model. The timing of the maximum surge at Campbell River was identical with observations and too early at the other the stations.

%\begin{table}[h]
%\centering 
%\begin{tabular}{|l |c c c c c | c c c c c|} 
%\hline 
%& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
%\hline 
%Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
%\hline 
%Point Atkinson &  9.8           &  3.42                & 3.43                 &   0.034    & 0.991  &  7.7           &  0.290                 & 0.247                  &  0.128     & 0.950 \\
%Victoria       &  9.1           &  2.16                & 2.11                 &   0.070    & 0.978  &  6.2           &  0.222                 & 0.202                  &  0.171     & 0.946 \\
%Patricia Bay   & 10.0           &  2.57                & 2.53                 &   0.049    & 0.985  &  6.6           &  0.235                 & 0.203                  &  0.144     & 0.947 \\
%Campbell River & 10.7           &  3.20                & 3.26                 &   0.037    & 0.989  &  6.2           &  0.254                 & 0.236                  &  0.140     & 0.952 \\
%\hline 
%\end{tabular}
%\caption{Evaluation of model performance for the Nov 2006 hindcast. Statistics were calculated over the period Nov 14, 2006 to Nov 17, 2006.}
%\label{tab:nov2006stat} 
%\end{table}

@@

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 28 cm higher than the modelled residual, however the timing of the surge agrees almost exactly as displayed in Figure \ref{fig:dec2006}. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 44 cm. At all of the locations, the model predicted the maximum surge to occur within 3 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb 4, 2006 case. All Willmott skill scores rank above 0.89 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The modelled residuals are on average too low and have a lower skill score than the Feb 2006 case. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

%\begin{table}[h]
%\centering 
%\begin{tabular}{|l |c c c c c | c c c c c|} 
%\hline 
%& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
%\hline 
%Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
%\hline 
%Point Atkinson & 10.5           &  3.45                & 3.47                 &   0.025    & 0.994  & 11.1           &  0.212                 & 0.186                  &  0.275     & 0.891 \\
%Victoria       &  9.6           &  2.23                & 2.16                 &   0.042    & 0.984  &  7.9           &  0.190                 & 0.156                  &  0.223     & 0.911 \\
%Patricia Bay   &  9.2           &  2.63                & 2.58                 &   0.031    & 0.990  &  9.5           &  0.193                 & 0.149                  &  0.260     & 0.889 \\
%Campbell River & 10.6           &  3.23                & 3.28                 &   0.030    & 0.992  &  9.4           &  0.191                 & 0.163                  &  0.247     & 0.905 \\
%\hline 
%\end{tabular}
%\caption{Evaluation of model performance for the Dec 2006 hindcast. Statistics were calculated over the period Dec 14, 2006 to Dec 17, 2006.}
%\label{tab:dec2006stat} 
%\end{table}

@@

The modelled water levels and residuals are too high for this hindcast, except at Victoria (Table \ref{tab:statistics}). Although Victoria has the smallest mean error in both the water level and residuals, its Willmott skill score and $\gamma^2$ values rank low when compared to the other stations. Generally, the skills scores are very good, scoring above 0.89 at all stations.

%\begin{table}[h]
%\centering 
%\begin{tabular}{|l |c c c c c | c c c c c|} 
%\hline 
%& \multicolumn{5}{|c|}{Water Level ($\eta$)}        & \multicolumn{5}{|c|}{Residual ($\delta$)} \\ 
%\hline 
%Location       & $\bar{e}$ (cm) & $\bar{\eta_{o}}$ (m) & $\bar{\eta_{m}}$ (m) & $\gamma^2$ & $WS$   & $\bar{e}$ (cm) & $\bar{\delta_{o}}$ (m) & $\bar{\delta_{m}}$ (m) & $\gamma^2$ & $WS$ \\
%\hline 
%Point Atkinson & 14.0           &  3.44                & 3.55                 &   0.012    & 0.995  &  6.1           &  0.313                 & 0.331                  &  0.237     & 0.914 \\
%Victoria       &  9.9           &  2.26                & 2.24                 &   0.023    & 0.994  &  4.8           &  0.323                 & 0.293                  &  0.245     & 0.899 \\
%Patricia Bay   & 11.2           &  2.60                & 2.65                 &   0.020    & 0.993  &  6.0           &  0.277                 & 0.290                  &  0.272     & 0.901 \\
%Campbell River & 13.5           &  3.22                & 3.27                 &   0.019    & 0.995  &  4.9           &  0.326                 & 0.314                  &  0.217     & 0.924 \\
%\hline 
%\end{tabular}
%\caption{Evaluation of model performance for the Nov 2009 hindcast. Statistics were generated over the period Nov 17, 2009 to Nov 22, 2009.}
%\label{tab:nov2009stat} 
%\end{table}
